---
title: "CD8_rnaseq"
author: "Sobur"
date: "2024-06-21"
output:
  pdf_document: default
  html_document: default
---

```{r}
if (!requireNamespace("BioManager", quitely = TRUE))
  install.packages("BioManager")

BiocManager::install("DESeq2")
```


```{r}
library(DESeq2)
library(tidyverse)
```

### Import Count data from data dirctory
```{r}
Counts <- read.delim("data/counts_matrix.csv", header = TRUE, row.names = 1, sep = ",")
```


### For the convenience of DESeq2 analysis, remove the GeneSymbol Column for now

```{r}
Counts <- select(Counts, -GeneSymbol)
```



###If we zoom in close to zero, we can see a large number of genes with counts of zero:
```{r}
ggplot(Counts) +
  geom_histogram(aes(x = Bl1), stat = "bin", bins = 200) + 
  xlim(-5, 1000)  +
  xlab("Raw expression counts") +
  ylab("Number of genes")
```

### Eliminating low-count or low-expressed genes is a crucial preprocessing step. This helps to reduce noise and improve the power of downstream analyses, such as differential expression analysis. The specific thresholds and rules for raw count elimination can vary depending on the dataset and the research objectives

```{r}
Counts <- Counts[rowSums(Counts) > 50, ]
```

```{r}
condition <- factor(c(rep("Blood", 4), rep("Fat", 2), rep("IEL", 3), rep("Kidney", 3),
                rep("Liver", 2), rep("SG", 3), rep("Spleen", 5)))
```

```{r}
coldata <- data.frame(row.names = colnames(Counts), condition)
```

```{r}
colnames(Counts) %in% rownames(coldata)
```


###Create a DESeqDataSet object (dds) that combines the count data and the sample metadata, and specifies the experimental design
###This function creates a DESeqDataSet object, which is the primary data structure used by DESeq2 for differential expression analysis.

```{r}
dds <- DESeqDataSetFromMatrix(countData = Counts,
                              colData = coldata,
                              design = ~ condition) # design is the column name in the coldata with condition or for here tissue type
```

### Initiate the comprehensive process of analyzing RNA-seq data to identify differentially expressed genes between the conditions specified in your experimental design. This step is crucial for obtaining biologically meaningful insights from  RNA-seq experiments.
- estimating size factors
- estimating dispersions
- gene-wise dispersion estimates
- mean-dispersion relationship
- final dispersion estimates
- fitting model and testing



```{r}
dds <- DESeq(dds)
```

The results function in DESeq2 is used to extract the results of the differential expression analysis from a DESeqDataSet object. It provides a table of statistics for each gene, including the log2 fold changes, p-values, and adjusted p-values.

```{r}
res <- results(dds)
res
```


```{r}
# Plot dispersions
plotDispEsts(dds, main="Dispersion plot")

```



###The vst function from the DESeq2 package applies a variance stabilizing transformation to the count data. This transformation aims to stabilize the variance across the range of mean values, making the data more suitable for downstream analyses such as clustering or visualization.
```{r}
vsdata <- vst(dds, blind = FALSE)
```

### QC on the data by creating a PCA plot

```{r}
#Basic PCA plot
plotPCA(vsdata, intgroup = "condition")
```
```{r}
pcaData <- plotPCA(vsdata, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
# Install ggalt package if not already installed
if (!requireNamespace("ggalt", quietly = TRUE)) {
  install.packages("ggalt")
}
```

```{r}
library(ggalt)
library(ggrepel)
```


```{r}
centers <- pcaData %>%
  group_by(condition) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2))
```

```{r}
# Function to get convex hull for each group
find_hull <- function(df) df[chull(df$PC1, df$PC2), ]

# Apply the function to each condition group
hulls <- pcaData %>%
  group_by(condition) %>%
  do(find_hull(.))
```

### Create a customized PCA plot like the figure

```{r}
PCA_plot <- ggplot(pcaData, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 2) +  # Points
  geom_polygon(data = hulls, aes(x = PC1, y = PC2, group = condition, fill = condition), alpha = 0.2) +  # Polygons around points
  geom_text_repel(data = centers, aes(x = PC1, y = PC2, label = condition), size = 5, show.legend = FALSE) +  # Condition labels
  xlab(paste0("PC1: ", percentVar[1], "% variance")) + 
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of VST Data") +
  labs(color = "Tissue", fill = "Tissue") +  # Label the legend as "Tissue"
  theme_classic() +
  theme(legend.position = "right")

png("output/PCA_plot_fig1c.png", res = 250, width = 1500, height = 1500)
print(PCA_plot)
dev.off()

```

```{r}
### Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)

### Extract the rlog matrix from the object
rld_mat <- assay(rld)    ## assay() is function from the "SummarizedExperiment" package that was loaded when you loaded DESeq2

### Compute pairwise correlation values
rld_cor <- cor(rld_mat)    ## cor() is a base R function
```

```{r}
library(pheatmap)
```


```{r}
### Plot heatmap
pheatmap(rld_cor)
```
```{r}
contrast_oe <- c("condition", "Blood", "Spleen")
```

```{r}
res_unsh <- results(dds, contrast=contrast_oe)
```

```{r}
plotMA(res_unsh, ylim=c(-2,2))
```



```{r}
library(ashr)
res_sh <- lfcShrink(dds, contrast=contrast_oe, res = res_unsh, type = "ashr")
```

```{r}
plotMA(res_sh, ylim=c(-2,2))
```

```{r}
sigs <- res_unsh

```
```{r}
sigs_tb <- sigs %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
```

```{r}
sigsOE <- sigs_tb 
```


```{r}
normalized_counts <- counts(dds, normalized=TRUE)
```


```{r}
normalized_counts <- normalized_counts %>% 
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()

```


```{r}
norm_OEsig <- normalized_counts %>% 
  filter(gene %in% sigsOE$gene) %>% 
  data.frame() %>%
  column_to_rownames(var = "gene") 
```


```{r}
library(pheatmap)
```


```{r}
pheatmap(norm_OEsig, 
         cluster_rows = T, 
         show_rownames = F,
         border_color = NA, 
         fontsize = 10, 
         scale = "row", 
         fontsize_row = 10, 
         height = 20)
```



#Clustering Heatmap
### Define contrasts comparing each tissue with spleen

```{r}
contrasts <- list(
  Blood_vs_Spleen = c("condition", "Blood", "Spleen"),
  Fat_vs_Spleen = c("condition", "IEL", "Spleen"),
  IEL_vs_Spleen = c("condition", "Kidney", "Spleen"),
  Kidney_vs_Spleen = c("condition", "SG", "Spleen"),
  Liver_vs_Spleen = c("condition", "Fat", "Spleen"),
  SG_vs_Spleen = c("condition", "Liver", "Spleen")
)
```

### For each contrast, extract DEGs based on the criteria log2(fold-change) > 1 and Padj < 0.05.
```{r}
# List to store DEGs for each contrast
all_degs <- list()

# Loop through each contrast
for (con in names(contrasts)) {
  # Get results for the contrast
  res <- results(dds, contrast = contrasts[[con]])
  
  # Filter DEGs based on log2 fold-change and adjusted p-value, excluding NA values
  de_genes <- res[abs(res$log2FoldChange) > 1 & !is.na(res$padj) & res$padj < 0.05, ]
  
  # Add filtered DEGs to the list
  all_degs[[con]] <- de_genes
}


```


```{r}
# Combine all DEGs into a single dataframe (or list)
all_degs_combined <- do.call(rbind, all_degs)

# Optionally, remove duplicates if needed
all_degs_combined <- unique(all_degs_combined)

```


```{r}
plotMA(all_degs_combined, ylim=c(-2,2))
```
```{r}
sigs_tb <- all_degs_combined %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()

#####Extracting significant differentially expressed genes

### Set thresholds
padj.cutoff <- 0.01
lfc.cutoff <- 1


#Now we can subset that table to only keep the significant genes using our pre-defined thresholds:
  
sigOE <- sigs_tb %>%
  filter(padj < padj.cutoff & abs(log2FoldChange) > lfc.cutoff)
```

#Visualizing the results
```{r}
# Install and load required packages
if (!requireNamespace("pheatmap", quietly = TRUE)) {
  install.packages("pheatmap")
}
library(pheatmap)

```

```{r}

library(DEGreport)
library(RColorBrewer)
library(pheatmap)

```


```{r}
#Let’s create tibble objects from the meta and normalized_counts data frames before we start plotting.
#This will enable us to use the tidyverse functionality more easily.

# Create tibbles including row names
mov10_meta <- coldata %>% 
  rownames_to_column(var="samplename") %>% 
  as_tibble()
```

### Extract the normalized count data
```{r}
normalized_counts <- counts(dds, normalized=TRUE)
```


```{r}
normalized_counts <- normalized_counts %>% 
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()
```

```{r}
### Extract normalized expression for significant genes from the OE and control samples (4:9), and set the gene column (1) to row names
norm_OEsig <- normalized_counts %>% 
  filter(gene %in% sigOE$gene) %>% 
  data.frame() %>%
  column_to_rownames(var = "gene") 
```

```{r}
desired_order_cols <- c("Spl", "Bl", "IEL", "Kid", "SG", "Fat", "Liv")
# Extract the prefix of each column name (e.g., Spl, Bl, etc.)
column_prefixes <- gsub("[0-9]", "", names(norm_OEsig))

# Match the prefix with the desired order
ordered_indices <- unlist(lapply(desired_order_cols, function(prefix) which(column_prefixes == prefix)))

# Reorder the columns of norm_OEsig
norm_OEsig_ordered <- norm_OEsig[, ordered_indices]

```


```{r}
### Annotate our heatmap (optional)
annotation <- mov10_meta %>% 
  select(samplename, condition) %>% 
  data.frame(row.names = "samplename")

```

```{r}
# Define the desired order of conditions
desired_order <- c("Spleen", "Blood", "IEL", "Kidney", "SG", "Fat", "Liver")

# Convert the 'condition' column to a factor with the specified order
annotation$condition <- factor(annotation$condition, levels = desired_order)

# Reorder the annotation data frame based on the 'condition' column
annotation <- annotation[order(annotation$condition), , drop = FALSE]
```



```{r}
# Extract conditions from colData (adjust as per your actual data structure)
conditions <- colData(dds)$condition

# Merge conditions with all_degs_combined based on row names or another identifier
all_degs_combined <- cbind(all_degs_combined, condition = conditions[rownames(all_degs_combined)])

```


```{r}
# Create a color palette for the heatmap
heatmap_colors <- colorRampPalette(c("blue3", "white","red3"))(500)

# Generate the heatmap
heatmap <- pheatmap(norm_OEsig_ordered, 
         cluster_rows = FALSE,
         cluster_cols = F, 
         show_rownames = FALSE,
         show_colnames = FALSE, # Remove column names from the bottom
         annotation_col = annotation, # Add annotation on the top
         border_color = NA, 
         fontsize = 10, 
         scale = "row", 
         fontsize_row = 10, 
         color = heatmap_colors) # Set the custom color palette

png("output/heatmap.png", res = 250, width = 2200, height = 3000)
print(heatmap)
dev.off()
```


